---
title: "POMA: Statistical analysis tool for targeted metabolomic data"
subtitle: "Exploratory Analysis"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    toc: yes
    theme: united
params:
  n: NA
---

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
data <- params$n
#data <- read_csv("/home/pol/Escritorio/POMA/ST000284/MET_CRC_ST000284.csv")
colnames(data)[1:2] <- c("ID", "Group")
```

# Know your data

  + Your data has **`r nrow(data)`** subjects, **`r ncol(data)-2`** metabolites and **`r length(table(data$Group))`** groups, that are **`r noquote(paste(shQuote(levels(as.factor(data$Group))), collapse=", "))`**.   

  + A **`r round(sum(is.na(data))/(nrow(data)*(ncol(data)-2))*100, 3)`%** of the values in your data are NAs (missing values).   

# Subject count by Groups

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
library(ggplot2)
library(plotly)
library(reshape)

ggplotly(ggplot(data, aes(Group, fill = Group)) + 
     geom_bar(stat = "count") +
     labs(title = "Frequency by Groups", x = "Group") +
     theme_minimal())
```

# Variable Distribution

## Boxplots

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
data2 <- melt(as.data.frame(data[,2:ncol(data)]))

data2%>%
  group_by(Group)%>%
  plot_ly(x=~variable, y= ~value, color= ~Group, legendgroup=~Group, type = "box") %>%
  layout(title = "Boxplot of all metabolites", y = "Concentration")

data2%>%
  group_by(Group)%>%
  plot_ly(x=~Group, y= ~value, color= ~Group, legendgroup=~Group, type = "box") %>%
  layout(title = "Boxplot of all metabolites by Group", y = "Concentration")

data2%>%
  group_by(Group)%>%
  plot_ly(x=~variable, y= ~log2(value), color= ~Group, legendgroup=~Group, type = "box") %>%
  layout(title = "Boxplot of all metabolites (log2)", y = "log2(Concentration)")

data2%>%
  group_by(Group)%>%
  plot_ly(x=~Group, y= ~log2(value), color= ~Group, legendgroup=~Group, type = "box") %>%
  layout(title = "Boxplot of all metabolites by Group (log2)", y = "log2(Concentration)")
```

## Check Normality and Homogeneity

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
normality <- data.frame(pval=apply(data[,3:ncol(data)],2,function(x) {shapiro.test(x)$p.value}))
normality$pval <- p.adjust(normality$pval, method = "fdr")
percent.norm <- round((length(normality[normality$pval > 0.05,])/nrow(normality))*100,3)

normality$names <- rownames(normality)
names.normality <- normality[normality$pval > 0.05,2]

##

homo <- data.frame(pval=apply(data[,3:ncol(data)],2,function(x) {bartlett.test(x, data$Group)$p.value}))
homo$pval <- p.adjust(homo$pval, method = "fdr")
percent.homo <- round((length(homo[homo$pval > 0.05,])/nrow(homo))*100,3)

homo$names <- rownames(homo)
names.homo <- homo[homo$pval > 0.05,2]
```

  + A **`r percent.norm`%** of variables in your dataset have a normal distribution before log transformation. This variables are **`r noquote(paste(shQuote(names.normality), collapse=", "))`**.   
  + A **`r percent.homo`%** of variables in your dataset have an homogenic distribution before log transformation. This variables are **`r noquote(paste(shQuote(names.homo), collapse=", "))`**.    

## Density plots

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
ggplotly(ggplot(data2, aes(value, colour = variable)) +
    geom_density() + 
    theme_minimal() + 
    labs(title = "Density plot for all Variables", x = "Concentration"))

ggplotly(ggplot(data2, aes(log2(value), colour = variable)) +
    geom_density() + 
    theme_minimal() + 
    labs(title = "Density plot for all Variables", x = "log2(Concentration)"))
```


# High Correlated Variables

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
mycorr <- cor(data[,3:ncol(data)])
mycorr <- which(mycorr > 0.97 & mycorr != 1, arr.ind=TRUE) 
num.corr <- nrow(mycorr)
```

There are **`r num.corr`** high correlated metabolites in your data. This metabolites are **`r noquote(paste(shQuote(rownames(mycorr)), collapse=", "))`**.

# PCA

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
X <- as.matrix(data[,3:ncol(data)])
Y <- as.factor(data$Group) 
pca.res2<-mixOmics::pca(X, ncomp = 3, center = T, scale = T)  
PCi<-data.frame(pca.res2$x,Groups=Y)
ggplotly(ggplot(PCi,aes(x=PC1,y=PC2,col=Groups))+
           geom_point(size=3,alpha=0.5) +
           scale_color_manual(values = c("#FF1BB3","#A7FF5B","#99554D","blue","darkgoldenrod2","gray9")) + 
           theme_minimal())
```

